The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2024-01-29 paper.tex
remove non-free AW3D indication from paper, remove ... in lists
remove old comments
removed two space symbols, TODOs at 'unnecessary draft sections'

%\paragraph{Skeleton}
The largest or most comprehensive skeletal anatomical structure, the human skeleton, is the framework composed of all the bones and teeth of a human being.
The TA~\citep{ta2} equates the skeleton with the \enquote{Systema skeletale (skeletal system)}, which refers to the passive part of the musculoskeletal system, including the articulated bony skeleton with teeth, as well as cartilage, ligaments, and other joints.
ANNO does not adopt this equivalence, and \enquote{Skeleton } here refers to the entirety of a human's bones and teeth.
However, since teeth constitute their own tissue, they are treated separately, allowing all other components of the skeletal system, such as cartilage or ligaments (as tissues) or joints (e.g. as specific aspects), to be integrated and referenced separately in ANNO.
The anthropologically meaningful distinction between \enquote{Skeleton} and \enquote{Systema skeletale} can be maintained, while still referencing both within the TA concept.

@@

%
%\textbf{UNFINISHED DRAFT SECTION, TO BE HEAVILY REWORKED AND CUT}
%\subsections{General considerations}
%All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
%
%Anatomical Landmarks (Bone Parts) (Selection Criteria, scope and process, properties)
%Osteometric Landmarks (Anatomical Points)
%Measurements (Anatomical line)
%Relevant spatial anatomical objects (e.g., lines) used for phenotype determination functions Measurements were defined in terms of anatomical points/objects that bound them: consist of references to sections, the start point, midpoint, and end, and short definitions.
%Phenotypes
%Development of the ontology is based on the following selection criteria:
%The selection criteria are as follows:
Given the overwhelming number of bone parts present in the cranium, the initial selection was narrowed down to a selection of representative and relevant parts.
Overall, however, the aim is to include those that are of anthropological relevance, i.e. that contribute to navigation, localization, and identification on the bone.
Those structures included in the definitions of others were also to be defined.
For bones that lie on the median sagittal plane (e.g., mandible or sternum), bilateral landmarks are defined, each with a side reference.
For osteometric landmarks, all those already established in the core literature are to be used.
Furthermore, those that are relevant for meaningful measurement distances and can be annotated in \aw{} should be used.
For the measurement distances, those should be selected that are established in the majority of the core literature as well as necessary for discriminant functions and functions for estimating height and weight.
%In addition, it also had to be integrable into \aw{}.
The functions chosen were those with diagnostic value.
These were discriminant functions for sex determination and regression functions for body height and body weight estimation.

@@

%\textbf{UNFINISHED DRAFT SECTION}

>>>>

%\paragraph{Skeleton}
The largest or most comprehensive skeletal anatomical structure, the human skeleton, is the framework composed of all the bones and teeth of a human being.
The TA~\citep{ta2} equates the skeleton with the \enquote{Systema skeletale (skeletal system)}, which refers to the passive part of the musculoskeletal system, including the articulated bony skeleton with teeth, as well as cartilage, ligaments, and other joints.
ANNO does not adopt this equivalence, and \enquote{Skeleton} here refers to the entirety of a human's bones and teeth.
However, since teeth constitute their own tissue, they are treated separately, allowing all other components of the skeletal system, such as cartilage or ligaments (as tissues) or joints (e.g. as specific aspects), to be integrated and referenced separately in ANNO.
The anthropologically meaningful distinction between \enquote{Skeleton} and \enquote{Systema skeletale} can be maintained, while still referencing both within the TA concept.

@@

%
%\textbf{UNFINISHED DRAFT SECTION, TO BE HEAVILY REWORKED AND CUT} <-- is this still todo?
%\subsections{General considerations}
%All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
%
%Anatomical Landmarks (Bone Parts) (Selection Criteria, scope and process, properties)
%Osteometric Landmarks (Anatomical Points)
%Measurements (Anatomical line)
%Relevant spatial anatomical objects (e.g., lines) used for phenotype determination functions Measurements were defined in terms of anatomical points/objects that bound them: consist of references to sections, the start point, midpoint, and end, and short definitions.
%Phenotypes
%Development of the ontology is based on the following selection criteria:
%The selection criteria are as follows:
\todo{is this still an unfinished draft section?}
Given the overwhelming number of bone parts present in the cranium, the initial selection was narrowed down to a selection of representative and relevant parts.
Overall, however, the aim is to include those that are of anthropological relevance, i.e. that contribute to navigation, localization, and identification on the bone.
Those structures included in the definitions of others were also to be defined.
For bones that lie on the median sagittal plane (e.g., mandible or sternum), bilateral landmarks are defined, each with a side reference.
For osteometric landmarks, all those already established in the core literature are to be used.
Furthermore, those that are relevant for meaningful measurement distances and can be annotated in \aw{} should be used.
For the measurement distances, those should be selected that are established in the majority of the core literature as well as necessary for discriminant functions and functions for estimating height and weight.
%In addition, it also had to be integrable into \aw{}.
The functions chosen were those with diagnostic value.
These were discriminant functions for sex determination and regression functions for body height and body weight estimation.

@@

%\textbf{UNFINISHED DRAFT SECTION}
\todo{is this still an unfinished draft section?}
